Supplementary material

### 2017-07-05 Manuscript/ms.tex
Cleaning up parameterization

Adult surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
However, we can borrow from the field of 'integrated population modeling' from wildlife ecology and use multiple sources of information to estimate mosquito demographic rates (CITATIONS).
Since mosquito survival is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e. reports of human cases), given an appropriate underlying mechanistic model.

In this paper, we propose a mechanistic transmission model that connects mosquito abundance and survival with dengue epidemiology and human case reports.
We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
Lastly, we explore the utility of these estimated transmission risks for guiding targeted control efforts relative to using mosquito abundance alone.

@@

The complete differential equation model is then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\frac{dV_N}{dt} & = r_t V_N \\
\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_S &= V_N - V_E - V_I,
\end{align}
where
\begin{align}
d_t & = \rho_{vt} \exp(-\nu_t)\\
\nu_{t} & = \alpha_{\nu} \nu_{t-1} + \theta_{\nu} (\nu_{t-52} - \alpha_{\nu} \nu_{t-53}) +\epsilon_{\nu t}\\
r_t &= \alpha_r r_{t-1} + \theta_r (r_{t-52} - \alpha_r r_{t-53}) + \epsilon_{rt},\\
\end{align}
and
\begin{align}
\epsilon_{rt} & \sim \text{Normal}(0, \sigma_r)\\
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma_{\nu}).
\end{align}

@@

Several of the parameters in this model are assumed to fixed and known. 
The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) are also fixed at literature values.
Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.

The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters controlling mean, phase, and amplitude of the emergence rate ($\alpha_0,\alpha_1,\alpha_2$), the regression and variance parameters for the mosquito death rate process ($\beta_0,\mathbf{\beta},\sigma$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).  
Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).

For any given values of these parameters, the mosquito process model should be flexible enough to produce a good fit to the case reports data.
As a result, we expect that these parameters will be informed primarily by the mosquito surveillance data (i.e. these parameters add degrees of freedom that let the mosquito process model fit the surveillance data better while maintaining a good fit to the case data).

\begin{table}[!ht]
\label{fixedparms}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$N$ & Population size & 327801 & Vitoria Census\\
$1/d$ & Human life-span & 76 years & Vitoria Census\\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
$1/\rho_{vt}$ & Extrinsic incubation period & $\exp \left(\exp(1.9 - 0.04 T_t) + \frac{1}{14}\right)$ & \cite{Chan2012}\\
$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
\hline
$\exp(\beta_0)$ & Mean vector death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
$\exp(\alpha_0)$ & Mean vector net emergence rate & ?? & \\
$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
$E_0$ & Proportion initially exposed & $8\times 10 ^ {-5}$ & \\
$I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & 
\end{tabular}
\end{center}
\end{adjustwidth}
\end{table}

\subsection*{Model validation}

We evaluate the model's predictive ability by fitting to the first 191 weeks of data (up to the trough following the fourth outbreak), and predicting case reports for the subsequent year.
In order to make these predictions, we project weather covariates into the future using weekly averages of the first 191 weeks.

@@

Questions:
\begin{itemize}
\item How wells does the model fit the case reports and mosquito surveillance data?
\item What do the estimated time series of demographic rates look like? How smooth are they? How strong is the periodic signal relative to noise?
\item How far/how well can we make predictions?
\item What do the estimated demographic rates tell us about efficient implementation of control?
\end{itemize}

>>>>

Adult \textit{Aedes aegypti} surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
However, we can borrow from the field of 'integrated population modeling' and combine multiple sources of information to estimate mosquito demographic rates (CITATIONS).
Since mosquito survival is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e. reports of human cases), given an appropriate underlying mechanistic model.

In this paper, we propose a mechanistic model to connect mosquito abundance and survival with dengue epidemiology.
We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
Lastly, we explore the utility of these estimated transmission risks for guiding targeted control efforts relative to using mosquito abundance alone.

@@

The complete differential equation model is then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\frac{dV_N}{dt} & = r_t V_N \\
\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_S &= V_N - V_E - V_I.
\end{align}

We model the weekly mosquito mortality rate, $d_{vt}$, relative to $\rho_{vt}$:
\begin{equation}
d_{vt} = \rho_{vt} \exp(-\mu - \nu_t).
\end{equation}
Assuming mosquito lifespan and latent periods are exponentially distributed (as the differential equation framework implicitly does), the probability that a newly exposed mosquito will survive until it becomes infectious is then
\begin{equation}
\psi_t = \frac{\rho_{vt}}{\rho_{vt} + d_{vt}} = \text{logit}^{-1}(\mu + \nu_t).
\end{equation}
We then model $\nu_t$ and the mosquito population growth rate, $r_t$, using a (Box-Jenkins?) seasonal auto-regressive model (CITATION):
\begin{align}
\nu_{t} & = \alpha_{\nu} \nu_{t-1} + \theta_{\nu} (\nu_{t-52} - \alpha_{\nu} \nu_{t-53}) +\epsilon_{\nu t}\\
r_t &= \alpha_r r_{t-1} + \theta_r (r_{t-52} - \alpha_r r_{t-53}) + \epsilon_{rt},\\
\end{align}
where
\begin{align}
\epsilon_{rt} & \sim \text{Normal}(0, \sigma_r)\\
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma_{\nu}).
\end{align}

@@

Several of the parameters in this model are assumed to be fixed and known. 
The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) are also fixed at literature values.
Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.

The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the latent mosquito process autoregressive ($\alpha_r$, $\alpha_{\nu}$, $\theta_r$, $\theta_{\nu}$) and variance ($\sigma_r$, $\sigma_{\nu}$) parameters, and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).  
Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).

\begin{table}[!ht]
\label{fixedparms}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$N$ & Population size & 327801 & Vitoria Census\\
$1/d$ & Human life-span & 76 years & Vitoria Census\\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
$1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
$\exp(\mu)$ & Mean ratio of mortality to incubation rates & 3.6 & \cite{Chan2012,Brady2013} \\
$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
\hline
$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
$E_0$ & Proportion initially exposed & $8\times 10 ^ {-5}$ & \\
$I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & 
\end{tabular}
\end{center}
\end{adjustwidth}
\end{table}

@@

Questions:
\begin{itemize}
\item How wells does the model fit the case reports and mosquito surveillance data?
\item What do the estimated time series of demographic rates look like? How smooth are they? How strong is the periodic signal relative to noise?
\item How do peaks in $\psi$ line up with peaks in $q_t$?
\item How effective is control informed by $\psi$ compared to control informed by $q_t$?
\end{itemize}
